Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-02 rmd/manuscript.Rmd
edit: manuscript

We ran `r max(df_sim$n_rep)` simulation replicates with `r n_sim` timesteps of trophic dynamics (including `r n_sim - unique(df_sim$n_timestep)` steps of warmup and burn-in). During the warmup period (`r unique(df_sim$n_warmup)` time steps), we introduced propagules into habitat patches every 10 time steps to allow species to establish populations. Propagules are random draws from $\text{Poisson}(\lambda)$ ($\lambda=100, 10,~\text{and}~1$ for $B$, $C$, and $P$, respectively). Subsequently, we ran a burn-in period with no propagules for `r unique(df_sim$n_burnin)` time steps to remove the effects of initial conditions. We obtained the global FCL $\bar{\tau}$ for the last `r unique(df_sim$n_timestep)` time steps. Note that simulation replicates are differ in ecosystem size ($N_{p}\sim \text{Unif}(10,150)$) and ecosystem complexity ($P_{b} \sim \text{Unif}(0.01, 0.99)$) so we can estimate their effects on FCL. All analysis was performed in R 4.1.0 [@rcoreteamLanguageEnvironmentStatistical2021]

@@

The exception is the "chain" scenario (i.e., no omnivory) with no disturbance. In this scenario, ecosystem size had a clear positive effect on FCL while ecosystem complexity had a negative influence. These opposing influences of ecosystem size and complexity may reflect their contrasting impacts on the total carrying capacity within a network (i.e., carrying capacity summed across all patches): ecosystem size increases the total carrying capacity whereas ecosystem complexity decreases it. However, these patterns characteristic to the "chain" scenario diminished when considering some levels of omnivory because food webs with omnivory were less reliant on the carrying capacity of the basal species. In addition, omnivory gave rise to more robustness/resilience to increased disturbance frequency.

@@

@@

The ecosystem size hypothesis of FCL has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012], although empirical work so far has rarely elaborated on the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, ecosystem complexity has not been incorporated in theoretical explorations of FCL to date. Here, we show that FCL responds positively to increasing ecosystem size and complexity, assessed as the number of patches and branching probability in river networks. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of FCL. However, the relationship between FCL and ecosystem complexity varied with other extrinsic and intrinsic factors, supporting the emerging paradigm that FCL is regulated by multiple factors [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other systems with branching structures, such as mountain ranges and cave systems. Our findings should also be relevant to non-branching systems, where different metrics of ecosystem complexity would be necessary.

While increased disturbance frequency consistently reduced FCL as predicted by the dynamical stability hypothesis [@pimmNumberTrophicLevels1977], more important findings in this study were evident from their synergistic effects with ecosystem structure. Our low disturbance regime scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity. Consequently, the food web structure largely controlled FCL. However, under high disturbance regimes, only large and complex networks could provide diverse and heterogeneous habitats with potential refugia, allowing species to thrive and maintain long food chains. While empirical studies rarely looked at the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

The interactive effect of disturbance and ecosystem structure has important implications for ecosystem management because anthropogenic climate change is predicted to increase the frequency and magnitude of disturbances [@brunnerFutureStreamflowRegime2020; @mittalImpactHumanIntervention2016]. River networks are especially vulnerable due to their propensity to go dry during droughts [@zipperPervasiveChangesStream2021] as well as susceptibility to more intense flood disturbance with increasing storm intensity [@schiermeierIncreasedFloodRisk2011]. Therefore, the preservation of habitat diversity conferred by ecosystem size and complexity is critical to biodiversity conservation under rapid environmental changes. In addition, maintaining habitat connectivity is particularly important for large, linearly-arranged networks. Although dispersal has received less attention in food chain research [but see @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012; @wangMetapopulationCapacityDetermines2021], our simulation highlighted food webs in these networks are more vulnerable to habitat fragmentation because habitat patches are inherently isolated from one another. Dams and road crossings have altered the connectivity between suitable habitat patches or have completely blocked dispersal [@ishiyamaPredictingEcologicalImpacts2018; @moritaEffectsHabitatFragmentation2002; @perkinFragmentationAltersStream2012]. Light pollution has also negatively affected the ability of adult aquatic insects to disperse between habitat patches [@perkinEffectsArtificialLighting2014]. These human activities may disproportionately impact food webs in large watersheds with simple network structures.

It is not surprising that FCL declined with increasing omnivory because it effectively compresses a food web (**Figure 1C**) and/or intensifies competition between consumer and predator species. However, how omnivory influences the associations with ecosystem structure has rarely been explored [@takimotoEffectsProductivityDisturbance2012; @wardMechanisticTheoryAquatic2017]. An important result in our simulation is that omnivory does not change the qualitative relationships between FCL and ecosystem properties, as opposed to productivity effects. This result provides a possible explanation for why the ecosystem size hypothesis has been consistently supported in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. In addition, our results imply that the size and complexity effects are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. However, such substantial changes in food web structure may have limited influences on how ecosystem structure dictates FCL.

>>>>

We ran `r max(df_sim$n_rep)` simulation replicates with `r n_sim` timesteps of trophic dynamics (including `r n_sim - unique(df_sim$n_timestep)` steps of warmup and burn-in). During the warmup period (`r unique(df_sim$n_warmup)` time steps), we introduced propagules into habitat patches every 10 time steps to allow species to establish populations. Propagules are random draws from $\text{Poisson}(\lambda)$ ($\lambda=100, 10,~\text{and}~1$ for $B$, $C$, and $P$, respectively). Subsequently, we ran a burn-in period with no propagules for `r unique(df_sim$n_burnin)` time steps to remove the effects of initial conditions. Note that simulation replicates are differ in ecosystem size ($N_{p}\sim \text{Unif}(10,150)$) and ecosystem complexity ($P_{b} \sim \text{Unif}(0.01, 0.99)$) so we can estimate their effects on FCL.

We obtained the global FCL $\bar{\tau}$ for the last `r unique(df_sim$n_timestep)` time steps. We also calculated the proportion of the patch state ("no species", "B", "B + C", "B + P", and "B + C + P") to infer mechanisms that drive changes in FCL. All analysis was performed in R 4.2.1 [@rcoreteamLanguageEnvironmentStatistical2021]

@@

The exception is the "chain" scenario (i.e., no omnivory) with no disturbance. In this scenario, ecosystem size had a clear positive effect on FCL while ecosystem complexity had a negative influence. These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increases the total carrying capacity whereas ecosystem complexity decreases it. However, these patterns characteristic to the "chain" scenario diminished when considering some levels of omnivory because food webs with omnivory were less reliant on the carrying capacity of the basal species.

@@

Productivity and dispersal had limited influences on the qualitative relationship between FCL and ecosystem geometry. However, they had important direct effects on FCL. Productivity influences were dependent on omnivory. Overall, greater productivity increased (chain & strong omnivory) or decreased FCL (weak omnivory), although the effect size varied with ecosystem geometry and disturbance levels. Dispersal distance had a positive effect on FCL, particularly in scenarios with frequent and intense disturbance.

## Underlying state changes

@@

The ecosystem size hypothesis of FCL has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012], although empirical work so far has rarely elaborated on the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, ecosystem complexity has not been incorporated in theoretical explorations of FCL to date. Here, we show that FCL responds positively to increasing ecosystem size and complexity, assessed as the number of patches and branching probability in river networks. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of FCL. However, the relationship between FCL and ecosystem complexity varied with other extrinsic and intrinsic factors, supporting the emerging paradigm that FCL is regulated by multiple factors [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other systems with non-branching structures. Ecosystems are intrinsically complex -- forest stands possess complex vertical structures, which may provide heterogeneous habitats with different levels of vulnerability to wind and/or fire disturbance. Topographic variation can also mediate spatial disturbance dynamics by providing local refugia in a landscape. An explicit consideration of ecosystem complexity may therefore increase our ability to predict FCL.

The most important finding in this study is the synergistic effects between ecosystem geometry and disturbance. Our no disturbance scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity (except the chain scenario). However, as the frequency of intense disturbance increases, small and/or simple networks could not maintain long food chains because habitats were too homogeneous with limited disturbance refugia. While empirical studies rarely looked at the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

The synergy between ecosystem geometry and disturbance has also an implication for why the dynamical stability hypothesis [@pimmNumberTrophicLevels1977] has been least supported by the literature. While increased disturbance frequency consistently reduced FCL, its magnitude depended strongly on ecosystem geometry. This observation should provide a possible explanation for the limited support for the dynamical stability hypothesis, because such an interaction has been rarely assessed in previous studies. Although FCL research often assumes that the effects of environmental controls are independent of one another, exploring potential interactions among them could provide useful insights into what controls FCL in nature.

Our results are crucial given that anthropogenic climate change is predicted to increase the frequency and magnitude of disturbances [@brunnerFutureStreamflowRegime2020; @mittalImpactHumanIntervention2016]. River networks are especially vulnerable due to their propensity to go dry during droughts [@zipperPervasiveChangesStream2021] as well as susceptibility to more intense flood disturbance with increasing storm intensity [@schiermeierIncreasedFloodRisk2011]. Therefore, the preservation of habitat diversity conferred by ecosystem size and complexity is critical to biodiversity conservation under rapid environmental changes. In addition, maintaining habitat connectivity is important because the effect of spatial refuges can emerge only through recolonization after disturbance. Dams and road crossings have altered the connectivity between suitable habitat patches or have completely blocked dispersal [@ishiyamaPredictingEcologicalImpacts2018; @moritaEffectsHabitatFragmentation2002; @perkinFragmentationAltersStream2012]. Light pollution has also negatively affected the ability of adult aquatic insects to disperse between habitat patches [@perkinEffectsArtificialLighting2014]. These human activities may increase the vulnerability of food webs to rapid environmental changes.

Another interesting result of our simulation is the interaction with omnivory. In particular, when the chain motif was combined with the no-disturbance scenario, ecosystem size and complexity had contrasting influences on FCL. It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity; therefore, the total carrying capacity within a network dictates FCL. The effect of ecosystem size is understandable because it simply increases the total carrying capacity. In contrast, ecosystem complexity, here assessed as branching probability, decreases the total carrying capacity as more branched networks have disproportionate numbers of headwater habitats with small local carrying capacities. However, the dependence of FCL on local carrying capacity diminished in weak and strong omnivory scenarios, explaining the lack of positive (ecosystem size) or negative FCL-geomtery relationship (ecosystem complexity) in these motifs with no disturbance.

Importantly, however, disturbance overwhelmed the effect of omnivory on the FCL-geometry relationships: as the frequency of intense disturbance increases, the positive relationships between FCL and ecosystem geometry emerged regardless of omnivory levels. This result provides a possible explanation for why the ecosystem size hypothesis has been consistently supported in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. Our results also imply that, if disturbance is strong and frequent, the effects of ecosystem geometry are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. However, such substantial food web reorganization may have limited influences on how ecosystem geometry dictates FCL.
